---
title: "Flight Data Analysis"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = FALSE, 
                      warning = FALSE,
                      comment = "")
```



# Student Information

**Name**: Sahith



# Introduction

A dataset of all domestic outbound flights from Dulles International Airport in 2016.

Airports depend on accurate flight departure and arrival estimates to maintain operations, profitability, customer satisfaction, and compliance with state and federal laws. Flight performance, including departure and arrival delays must be monitored, submitted to the Federal Aviation Agency (FAA) on a regular basis, and minimized to maintain airport operations. **The FAA considered a flight to be delayed if it has an arrival delay of at least 15 minutes**.

To perform an exploratory data analysis on all domestic flights from 2016 and produce an executive summary of key insights and recommendations.

The following is the airline flight terminology to familiarize with the industry: [Airline Flight Terms](https://www.myairlineflight.com)



## Dulles Flights Data

The `flights_df` data frame is loaded below and consists of 33,433 flights from IAD (Dulles International) in 2016. The rows in this data frame represent a single flight with all of the associated features that are displayed in the table below.

**Note**: If you have not installed the `tidyverse` package, please do so by going to the `Packages` tab in the lower right section of RStudio, select the `Install` button and type `tidyverse` into the prompt. If you cannot load the data, then try downloading the latest version of R (at least 4.0). The `readRDS()` function has different behavior in older versions of `R` and may cause loading issues.



```{r}

library(tidyverse)

flights_df <- readRDS(url('https://gmubusinessanalytics.netlify.app/data/dulles_flights.rds'))

```


### Raw Data

```{r}

flights_df

```


# Exploratory Data Analysis


Executives at this company have hired you as a data science consultant to evaluate their flight data and make recommendations on flight operations and strategies for minimizing flight delays.

You must think of **at least 8 relevant questions** that will provide evidence for your recommendations.

The goal of your analysis should be discovering which variables drive the differences between flights that are early/on-time vs. flights that are delayed.

Some of the many questions you can explore include:

-   Are flight delays affected by taxi-out time? Do certain airlines or time of 
    year lead to greater taxi out times (i.e. traffic jams on the runways)?

-   Are certain times of the day or year problematic?

-   Are certain destination or airlines prone to delays?


You must answer each question and provide supporting data summaries with either a summary data frame (using `dplyr`/`tidyr`) or a plot (using `ggplot`) or both.

In total, you must have a minimum of 5 plots and 4 summary data frames for the exploratory data analysis section. Among the plots you produce, you must have at least 4 different types (ex. box plot, bar chart,
histogram, heat map, etc...)

Each question must be answered with supporting evidence from your tables and plots.


## Question 1



**Question**: Are certain destination or airlines prone to delays?



**Answer**: Yes. According to the analysis, the United Airlines is more prone to delay (where it got delayed 3115 times) and when destination airport is considered, the Los Angeles airport is more prone to delay (where 730 flights got delayed). And, when Airlines and the destination airport it is going to is considered, United Airlines going to the San Francisco airport tops the list where it got delayed 541 times. But, from the boxplot it can be seen that if the two values for Unied Airlines (541 - going to San Francisco and 433 - going to Los Angeles) are considered as outliers, then the flight by United Airlines going to Denver is the one which got delayed most number of times (298).



To add additional R code chunks for your work, select `Insert` then `R` from the top of this notebook file.

```{r}

airlines_delay <- flights_df %>% 
  filter(arrival_delay >= 15) %>%
  group_by(airline) %>% 
  summarise(n_flights_delayed = n()) %>%
  arrange(desc(n_flights_delayed))

airlines_delay

```

```{r}

destination_delay <- flights_df %>%
  filter(arrival_delay >= 15) %>%
  group_by(dest_airport_name) %>%
  summarise(flights_delayed = n()) %>%
  arrange(desc(flights_delayed))

destination_delay

```

```{r}

delay_flights_dest <- flights_df %>% 
  filter(arrival_delay >= 15) %>%
  group_by(airline, dest_airport_name) %>% 
  summarise(delayed_flights = n()) %>%
  arrange(desc(delayed_flights))

delay_flights_dest

```

```{r}

ggplot(data = delay_flights_dest, 
       mapping = aes(x = delayed_flights, 
                     y = reorder(airline, delayed_flights))) +
  geom_boxplot(fill = "#006EA1") + 
  labs(title = "Boxplot of no. of flights got delayed by airlines",
       x = "No. of flights delayed", 
       y = "Airlines")

```


## Question 2



**Question**: Are certain times of the day or year problematic?



**Answer**: Yes. Based on the analysis done, July and June tops the list in which most number of flights got delayed (748 and 682 respectively). It may be due to the summer season where people travel more as this being a holiday season. Also, bad weather conditions like summer thunderstorms and fog sometimes can affect the flight journey.




```{r}

delay_flights_month <- flights_df %>%
  filter(arrival_delay >= 15) %>%
  group_by(month) %>%
  summarise(delayed_flights_mnth = n()) %>%
  arrange(desc(delayed_flights_mnth))

delay_flights_month

```

```{r}

ggplot(data = delay_flights_month, mapping = aes(x = reorder(month, -delayed_flights_mnth), 
                                                 y = delayed_flights_mnth)) + 
  geom_bar(stat = "identity", fill = "#006EA1", color = "white") + 
  labs(title = "Flights delayed in each month", 
       x = "Month", 
       y = "No. of flights delayed") +
  scale_x_discrete(labels = c("January" = "Jan", "February" = "Feb", "March" = "Mar", "April" = "Apr",
                              "May" = "May", "June" = "Jun", "July" = "Jul", "August" = "Aug",
                              "September" = "Sept", "October" = "Oct", "November" = "Nov",
                              "December" = "Dec"))

```


## Question 3



**Question**: Are flight delays of a particular airlines affected by taxi-out time?



**Answer**: Yes. Totaling the values of taxi_out for a particular airline gives the total time taken by the airline (which is the total sum of observations given in the data) to go to the runway from the gate, and comparing this to the total no. of flights delayed of that particular airline, it can be said that the taxi_out time has affect on the flight delays. 




```{r}

delay_flights_to <- flights_df %>%
  filter(arrival_delay >= 15) %>%
  group_by(airline) %>%
  summarise(sum_to = sum(taxi_out),
            avg_delay = mean(arrival_delay),
            delay_count = n()) %>%
  arrange(desc(sum_to))

delay_flights_to

```

```{r}

ggplot(data = delay_flights_to, mapping = aes(x = sum_to, y = delay_count, color = airline)) +
  geom_point(size = 2, alpha = 0.75) +
  labs(title = "Taxi_out vs delay_count by airline", 
       x = "Sum of taxi_out", 
       y = "No. of flights delayed")

```



## Question 4



**Question**: Do certain time of year lead to greater taxi out times (i.e., traffic jams on the runways)?



**Answer**: From the analysis, the taxi out time spent by an airline is more in June and July. Also, the average delay in these months is high when compared with other months. So, from this it can be observed that the summer season in the year leads to greater taxi out times.




```{r}

delay_month_high_to <- flights_df %>%
  filter(arrival_delay >= 15) %>%
  group_by(month) %>%
  summarise(max_to = max(taxi_out),
            avg_delay = mean(arrival_delay),
            delay_count = n()) %>%
  arrange(desc(max_to))

delay_month_high_to

```

```{r}

ggplot(data = delay_month_high_to, mapping = aes(x = reorder(month, -max_to), 
                                                 y = max_to)) + 
  geom_bar(stat = "identity", fill = "#006EA1", color = "white") + 
  labs(title = "Taxi out vs month", 
       x = "Month", 
       y = "Max of taxi_out") +
  scale_x_discrete(labels = c("January" = "Jan", "February" = "Feb", "March" = "Mar", "April" = "Apr",
                              "May" = "May", "June" = "Jun", "July" = "Jul", "August" = "Aug",
                              "September" = "Sept", "October" = "Oct", "November" = "Nov",
                              "December" = "Dec"))

```



## Question 5



**Question**: How Departure delays and arrival delays by certain airlines are related?



**Answer**: From this analysis, it can be observed that even though the number of flights delayed at departure are more for each airline, the number of flights delayed at arrival are less. 



```{r}

dept_delay  <- flights_df %>%
  filter(dep_delay > 0) %>%
  group_by(airline) %>%
  summarise(delay_dept = n()) %>%
  arrange(desc(delay_dept))

dept_delay

```

```{r}

arr_dept_airline <- cbind(dept_delay, airlines_delay)
arr_dept <- arr_dept_airline %>% select(-3)
arr_dept

```

```{r}

ggplot(data = arr_dept, mapping = aes(x = delay_dept, y = n_flights_delayed)) +
  geom_line(color = "#0072B2") + 
  geom_point(color = "#0072B2") +
  labs(title = "Departure delay vs arrival delay",
            x = "departure delay",
            y = "arrival delay")

```



## Question 6



**Question**: Does the weekday has any effect on delays and taxi in time?



**Answer**: Yes. Thursday, monday and friday are the top 3 days with most number of delayed flights. It may be due to the weekend getaway. 




```{r}

weekday_delay <- flights_df %>%
  filter(arrival_delay >= 15) %>%
  group_by(weekday) %>%
  summarise(avg_ti = mean(taxi_in),
    delay_count = n()) %>%
  arrange(desc(delay_count))

weekday_delay

```

```{r}

ggplot(data = weekday_delay, mapping = aes(x = reorder(weekday, -delay_count), 
                                                 y = delay_count)) + 
  geom_bar(stat = "identity", fill = "#006EA1", color = "white") + 
  labs(title = "Weekday vs delays", 
       x = "Day", 
       y = "No. of delays")

```


## Question 7



**Question**: Which flight number got delayed the most?



**Answer**: As said earlier, United airlines got delayed the most number of times. The flight number 1972 of United airlines is the one which got delayed many times.




```{r}

flightnum_delay <- flights_df %>%
  filter(arrival_delay >= 15) %>%
  group_by(flight_num, airline) %>%
  summarise(delay_count = n()) %>%
  arrange(desc(delay_count))

flightnum_delay

```



## Question 8



**Question**: Does the destination airport region has any effect on flights delayed?



**Answer**: Yes. Flights heading to West are getting delayed most number of times. 




```{r}

dest_regn <- flights_df %>%
  filter(arrival_delay >= 15) %>%
  group_by(dest_airport_region) %>%
  summarise(delay_count = n()) %>%
  arrange(desc(delay_count))

dest_regn

```

```{r}

ggplot(data = dest_regn, mapping = aes(x = reorder(dest_airport_region, -delay_count), 
                                                 y = delay_count)) + 
  geom_bar(stat = "identity", fill = "#006EA1", color = "white") + 
  labs(title = "Region vs delays", 
       x = "Region", 
       y = "No. of delays")

```



# Summary of Results

Write an executive summary of your overall findings and recommendations to the executives at Dulles Airport. Think of this section as your closing remarks of a presentation, where you summarize your key findings and make recommendations on flight operations and strategies for minimizing flight delays.

Your executive summary must be written in a [professional tone](https://www.universalclass.com/articles/writing/business-writing/appropriate-tone-in-business-communications.htm), with minimal grammatical errors, and should include the following
sections:

1.  An introduction where you explain the business problem and goals of your data analysis

    -   What problem(s) is this company trying to solve? Why are they important
        to their future success?

    -   What was the goal of your analysis? What questions were you 
        trying to answer and why do they matter?


2.  Highlights and key findings from your Exploratory Data Analysis section

    -   What were the interesting findings from your analysis and
        **why are they important for the business**?

    -   This section is meant to **establish the need for your recommendations** 
        in the following section
        

3.  Your recommendations to the company

    -   Each recommendation must be supported by your data analysis results

    -   You must clearly explain **why** you are making each recommendation and which results 
        from your data analysis support this recommendation

    -   You must also describe the potential business impact of your recommendation:

        -   Why is this a good recommendation?

        -   What benefits will the business achieve?
      


## Executive Summary

Please write your executive summary below. If you prefer, you can type your summary in a text editor, such as Microsoft Word, and paste your final text here.


> Airports must accurately monitor flight departure and arrival estimates in order to maintain good operations, profitability and customer satisfaction. In the given dataset, there are many causes which can alter the estimates of flight arrivals and departures and further which effects the airport operations and profitability. It should focus more on these factors so that the operations run smoothly and efficiently.

> The goal of my analysis is to analyse which factors are affecting the flight delays so that the airport staff can be prepared in advanced to resolve these issues. 

> The questions which I tried to answer in my analysis are:

    >> Are certain destination or airlines prone to delays?
    >> Are certain times of the day or year problematic?
    >> Are flight delays of a particular airlines affected by taxi-out time?
    >> Do certain time of year lead to greater taxi out times (i.e., traffic jams on the runways)?
    >> How Departure delays and arrival delays by certain airlines are related?
    >> Does the weekday has any effect on delays and taxi in time?
    >> Which flight number got delayed the most?
    >> Does the destination airport region has any effect on flights delayed?
    
    >> These are important to know because the airports can further process these issues and increase their profitability and customer satisfaction.
    
> Key findings from my analysis are:

    >> The United Airlines is the one which is more prone to delays
    >> The San Francisco airport to which flights are heading to is the one where more number of flights are getting delayed.
    >> Flights are getting delayed in the summer season of the year.
    
> Recommendations to the company:

    >> Company should more focus on United airlines traffic to reduce it's arrival delay in every destination airport. Question 1 will justify this recommendation.
    
    >> Summer season is the time of the year where most number of flights are getting delayed and also taxi out time is more. Company should focus on this by adding more itineraries to the destinations where more people are going. As this is the vacation season, most people visit tourist places. So, there's a need to resolve this issue to benefit customers. Question 2 and 4 will justify this recommendation.
    
    >> The taxi out time should be reduced so that the flights can reach destination on time. Question 3 will justify this recommendation.
    
    >> Flights are getting delayed on weekends like Thursdays, Mondays and Fridays. These might be the days most people travel. On these days the company should focus more on the airline traffic to benefit customers and increase the airports profitability without loosing the customer satisfaction. Question 6 will justify this recommendation.
    
    >> Flights heading to West region are more likely to get delayed. Company should rectify the cause and resolve this in order to maintain smooth operations. Question 8 will justify this recommendation.











